Efficient FPGA Implementation of\\ Overclocking Friendly Online Arithmetic Operators

### 2014-05-08 FPT2014/bare_conf.tex
Multiplier

\usepackage[cmex10]{amsmath}

\usepackage{algorithm}
\usepackage{algorithmic}

@@

\section{Background and Relevant Work}
\subsection{Online Arithmetic}

@@

\section{Digit-parallel Online Adder on FPGAs}
\subsection{Related Works}
There has been previous works focusing on efficient FPGA implementation of the digit-parallel online adder. Generally the existing approaches can be classified into three types:%, as discussed below. A more detailed review of current works can be found in xxx.
\begin{enumerate}
	\item efficient mapping of the digit-parallel online adder onto sophisticated FPGAs resources [xxx];
	\item multiple operands addition by designing compressor trees based on bit counters [xxx];
	\item modifying existing FPGA architecture for more efficient online addition [xxx] and for specific applications [xxx].
\end{enumerate}

@@

\begin{figure}[htbp]
	\centering
	\includegraphics[width=.5\textwidth]{./Figures/SDadder_FPGA_New.pdf}
	\caption{Modified structure of online adder. Left: an example of 3-digit online adder. The shaded part refers to the 2 logic blocks (LBs) that can be mapped onto 1 slice. Right: one LB. the dotted box outlines the logic that can be mapped onto 1 LUT and the corresponding fast-carry logic.}
	\label{Fig:ModifiedOA}
\end{figure}

@@

\begin{figure}[htbp]
	\centering
	\includegraphics[width=.48\textwidth]{./Figures/SDadder_FastCarry_New.pdf}
	\caption{Implementation of 2 logic blocks (LBs) in 1 FPGA slice which contains four 6-LUTs. NC stands for Not Care.}
	\label{Fig:SliceNew}
\end{figure}

\section{Digit-parallel Online Multiplier on FPGAs}

>>>>

\usepackage[cmex10]{amsmath}
\usepackage{amssymb}

\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{indentfirst}
\renewcommand{\algorithmicrequire}{ \textbf{Initialization:}}
\renewcommand{\algorithmicensure}{ \textbf{Recurrence:}}

@@

\section{Background: Online Arithmetic}
\subsection{Key Features of Online Arithmetic}
In conventional arithmetic, results are generated either from the least significant digit, e.g. addition and multiplication, or from the most significant digit, e.g. division and square root. This inconsistency in computing directions can result in large latency when propagating data among different operations. Online arithmetic was proposed to solve this problem [xxx]. With online arithmetic, both the inputs and the outputs are processed in a MSD-first manner. This enables parallelism and duplication among various operations, and the overall computation latency can be significantly reduced.

Online arithmetic was originally designed for digit-serial operation, of which the data flow is illustrated in Fig~\ref{Fig:OnlineDataFlow}. It can be seen that in order to generate the first output digit, $\delta$ digits of inputs are required and $\delta$ is called ``online delay''. Notice that $\delta$ is normally a constant, which is independent of the precision in a given operation. For ease of discussion, in the following of this paper the input data are normalized to fixed point numbers in the range $(-1,1)$. Based on this premise, the online representation of $N$-digit operands and result at iteration $j$ are given by (\ref{Eq:Online_Operands}), where $j\in[-\delta,N-1]$ and $r$ denotes the radix \cite{Ercegovac_Book}.

\begin{figure}[htbp]
  \centering
  %\vspace{-2.5ex}
  \includegraphics[width=.48\textwidth]{./Figures/OnlineArithmetic_DataFlow.pdf}
  %\vspace{-3ex}
  \caption{Dataflow in digit-serial Online Arithmetic. $\delta$ denotes the online delay.}
  %\vspace{-.5ex}
  \label{Fig:OnlineDataFlow}
\end{figure}

\begin{eqnarray}\label{Eq:Online_Operands}
\footnotesize
  X_{[j]}=\sum_{i=1}^{j+\delta}x_ir^{-i},~Y_{[j]}=\sum_{i=1}^{j+\delta}y_ir^{-i},~Z_{[j]}=\sum_{i=1}^{j}z_ir^{-i}
\normalsize
\end{eqnarray}

MSD-first operation is possible with the employment of the redundant number system \cite{RedundantNumber}, in which each digit is represented with a redundant digit set $\{-a, \cdots,-1,0, 1, \cdots, a\}$, where $a\in[r/2,r-1]$. Due to the redundancy, the MSDs of the result can be calculated only based on partial information of the inputs and then the value of the number can be revised by the following digits.

\subsection{Binary Online Addition}
Adders serve as a critical building block for arithmetic operations. To perform digit-parallel online addition, a redundant adder can be directly utilized. The adder structure diagram is shown in Fig~\ref{Fig:Radix2SD_adder}. A major advantage of the redundant number system over the standard ripple-carry based arithmetic is that the propagation of carry is eliminated, resulting in a precision-independent computation time for addition. As seen in Figure~\ref{Fig:Radix2SD_adder}, the computation delay of this adder is only 2 full adder (FA) delays for any operand word-length, with the cost of one extra FA for each digit of operands. This makes the online adder suitable for building up more complex arithmetic operators such as multipliers to accelerate the sum of partial products \cite{RedundantMult_1987}.

\begin{figure}[htbp]
	\centering
	\includegraphics[width=.5\textwidth]{./Figures/SDAdder.pdf}
%\vspace{-8.5ex}
	\caption{An $N$-digit binary digit-parallel online adder with $(N+1)$-digit outputs.}
	\label{Fig:Radix2SD_adder}
\end{figure}

In Section~\ref{Sec:OA_FPGA}, we will present a novel method of mapping the digit parallel online adder efficiently on FPGAs using the built-in carry resources.

\subsection{Binary Online Multiplication}
Multiplication is another key arithmetic operator. Typically the online multiplication is performed in a recursive digit-serial manner, as illustrated in Algorithm~\ref{Algorithm:OnlineMult} \cite{Ercegovac_OnlineMult} where both inputs and outputs are $N$-digit number as given in~(\ref{Eq:Online_Operands}). For a given iteration $j$, the product digit $z_j$ is generated through a selection function $sel()$. For the radix $r$ and a chosen digit set, there exits an appropriate selection method and a value of $\delta$ which ensure convergence \cite{Ercegovac_OnlineMult}. As radix-2 is used most commonly in computer arithmetic, we keep $r=2$ throughout this paper with the corresponding redundant digit set $\{\overline{1},0,1\}$. In this case $sel()$ is given by (\ref{Eq:SelFunc_OM}) \cite{Oregon_OnlineNetwork}. Notice that the election is made only based on 1 integer bit and 1 fractional bit of $W_{[j]}$.

\begin{algorithm}[tbp]
  \caption{Online Multiplication}
  \begin{algorithmic}[1]
    \REQUIRE~$X_{[-\delta]}=Y_{[-\delta]}=P_{[-\delta]}=0$
    \ENSURE~$for~~ j=-\delta,~-\delta+1,~\cdots,~N-1 ~~do$
      \begin{eqnarray}\label{Eq:OnlineMult_General}
        \begin{matrix}
          H_{[j]}   & = & r^{-\delta}\left(x_{j+\delta+1}\cdot Y_{[j+1]}+y_{j+\delta+1}\cdot X_{[j]}\right)\\
          W_{[j]}   & = & P_{[j]} + H_{[j]}\\
          z_j       & = & sel(W_{[j]})\\
          P_{[j+1]} & = & r\left(W_{[j]}-z_j\right)
        \end{matrix}
      \end{eqnarray}
  \label{Algorithm:OnlineMult}
  \vspace{-2ex}
  \end{algorithmic}
\end{algorithm}

\begin{eqnarray}\label{Eq:SelFunc_OM}
%\small
  sel(W_{[j]})=\begin{cases}
    1 & \text{ if } W_{[j]} \geqslant \frac{1}{2} \\
    0 & \text{ if } -\frac{1}{2}\leqslant W_{[j]}<\frac{1}{2} \\
    \overline{1} & \text{ if } W_{[j]}<-\frac{1}{2}
  \end{cases}
%\normalsize
\end{eqnarray}

In Section~\ref{Sec:OM_FPGA}, we will describe a modified online multiplication algorithm and its FPGA implementation, which are designed specifically targeting on digit parallel operations.

@@

\section{Digit Parallel Online Adder on FPGAs}\label{Sec:OA_FPGA}
\subsection{Related Works}
There has been previous works focusing on efficient FPGA implementation of the digit-parallel online adder. Generally the existing approaches can be classified into three types:%, as discussed below. A more detailed review of current works can be found in xxx.
\begin{enumerate}
	\item efficient mapping of the digit-parallel online adder onto sophisticated FPGAs resources [xxx];
	\item multiple operands addition by designing compressor trees based on bit counters [xxx];
	\item modifying existing FPGA architecture for more efficient online addition [xxx] and for specific applications [xxx].
\end{enumerate}

@@

\begin{figure}[tbp]
	\centering
	\includegraphics[width=.5\textwidth]{./Figures/SDadder_FPGA_New.pdf}
	\caption{Modified structure of online adder. Left: an example of 3-digit online adder. The shaded part refers to the 2 logic blocks (LBs) that can be mapped onto 1 slice. Right: one LB. the dotted box outlines the logic that can be mapped onto 1 LUT and the corresponding fast-carry logic.}
	\label{Fig:ModifiedOA}
\end{figure}

@@

\begin{figure}[tbp]
	\centering
	\includegraphics[width=.48\textwidth]{./Figures/SDadder_FastCarry_New.pdf}
	\caption{Implementation of 2 logic blocks (LBs) in 1 FPGA slice which contains four 6-LUTs. NC stands for ``Not Care''.}
	\label{Fig:SliceNew}
\end{figure}

\subsection{Performance Analysis}

\section{Digit Parallel Online Multiplier on FPGAs}\label{Sec:OM_FPGA}
\subsection{Algorithm of Digit Parallel Online Multiplication}
Algorithm xxx as described in Section.xxx can be synthesized into a unrolled digit parallel structure, which has been demonstrated to be ``overclocking friendly'' because timing errors initially affect the least significant digits [xxx]. This synthesis method is straightforward and can be utilized to design other digit parallel online operators.

However, in order to implement the digit parallel online multiplier (OM), large area budget is required because of the large area overhead of a single iteration stage and the overall stage numbers required. For instance in an $N$-digit OM, totally $(2N+\delta)$ iterations are required to generate $2N$ digits outputs, and $(N+\delta)$ iterations for the most significant $N$ digits result. Each iteration stage consists of all operations in Eq.xxx. In comparison, in an $N$-digit array multiplier there are only $(N-1)$ rows, each of which is basically an $N$-digit ripple-carry adder.

Instead of simply implementing the unrolled format of Algorithm xxx, it can be optimized given that all inputs are digit parallel. Initially the online delay $\delta$ is employed to generate the MSD of the results only based on partial information of the inputs. Nevertheless if all digits of inputs are given simultaneously, $\delta$ is no longer necessary. Due to the same reason, $H[j]$, which takes 1 digit of input per stage as shown in Algorithm xxx, can be optimized to take one partial product $Xy_j$ or $Yx_j$ per stage. The optimized digit parallel online multiplication algorithm is described in Algorithm xxx.

\subsection{FPGA Implementation}
The structure diagram of a 4-digit OM using the proposed algorithm is shown in Fig.~\ref{Fig:PMStructure}(a). Generally in an $N$-digit OM, there are only $N$ stages to generate results with $2N$ digits. Each stage can be efficiently implemented using FPGAs, and the structure diagram of stage $j$ is shown in Fig.~\ref{Fig:PMStructure}(b). In each stage, an online adder is used to derive $W_{[j]}$. The selection logic, as described in Section xxx and Eq.xxx, takes 3 digits input (6 bits) and generate 1 digit output (2 bits). Hence it can be implemented using two 6-LUTs for each output bit. Similarly the generation of $P_{[j]}$ can also be implemented using one 6-LUT, since only the integer parts of $W_{[j]}$ might change due to the selection of $z_j$. In addition, the structure of Stage 1 can be further optimized by removing the online adder, because $P_{[0]}=0$.

\begin{figure}[tbp]
	\centering
	\includegraphics[width=.5\textwidth]{./Figures/ParallelMult_Structure.pdf}
	\caption{(a) Structural diagram of a 4-digit online multiplier using the proposed algorithm. (b) Structure of one stage. The word-length of all signals are labelled in terms of the number of digits. $N$ denotes the word-length of the input signal.}
	\label{Fig:PMStructure}
\end{figure}

\subsection{Structure Optimization for the MSD Half of the Results}

Normally the OM is connected with other arithmetic operators in real applications. If the outputs of an OM is utilized for subsequent operations, only the most significant half of the products are required to maintain the consistency of word-length between inputs and outputs. In the conventional multiplier with standard binary arithmetic, this is achieved by either truncating or rounding the least significant half of the products. However both the computation time and the structure remains unchanged, because the results are generated from LSDs.

In comparison, the online multiplier offers the possibility to optimize the structure to a further extent if only the MSD half of the results are required, as the results are generated initially from the MSD. The modified structure diagram is illustrated in Fig.~\ref{Fig:PMStructure_MSDhalf}(a). The word-length of the online adder in a given Stage $j$ can be reduced, as shown in Fig.~\ref{Fig:PMStructure_MSDhalf}(b). This is because there is no carry propagation in the online adder, and all digits of the online adder are obtained in parallel.

\begin{figure}[tbp]
	\centering
	\includegraphics[width=.5\textwidth]{./Figures/ParallelMult_MSDhalf.pdf}
	\caption{(a) Modified structure of a 4-digit online multiplier which only generates the most significant 4-digit result. }
	\label{Fig:PMStructure_MSDhalf}
\end{figure}

\subsection{Performance Analysis}
